---
title: "Model Comparison: Communication Apprehension and Public Transportation Use"
author: "Yuhao Wang"
date: "2026.7.30"
output: 
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(broom)
library(ggplot2)
cleaned_data <- read_csv(here("clean_data", "cleaned_prca_data.csv"))
```

# Research Or Analytical Question

Which of the following models gives the lowest RMSE for predicting public transportation use frequency?

1. PTRS = b0 + b1*CA12 + e

2. PTRS = b0 + b1*CAg + b2*CAi + e  

3. PTRS = b0 + b1*CAg + b2*CAi + b3*(CAg * CAi) + e

Where:
- PTRS = Public Transportation Use Frequency (public_trans_freq)
- CA12 = Total Communication Apprehension (prca_total, 12 items)
- CAg = Group Communication Apprehension (Q1-Q6)
- CAi = Interpersonal Communication Apprehension (Q13-Q18)

# Answer, Response, + Summary of Results

## Sample Overview

```{r sample-info}
sample_info <- cleaned_data %>%
  summarise(
    total_n = n(),
    missing_ptrs = sum(is.na(public_trans_freq)),
    missing_ca12 = sum(is.na(prca_total)),
    missing_cag = sum(is.na(CAg)),
    missing_cai = sum(is.na(CAi)),
    complete_cases = sum(!is.na(public_trans_freq) & !is.na(prca_total) & !is.na(CAg) & !is.na(CAi))
  )
```

- **Sample Size:** `r sample_info$total_n` participants
- **Missing PTRS:** `r sample_info$missing_ptrs`
- **Missing CA12:** `r sample_info$missing_ca12`
- **Missing CAg:** `r sample_info$missing_cag`
- **Missing CAi:** `r sample_info$missing_cai`
- **Complete Cases:** `r sample_info$complete_cases` (`r round(sample_info$complete_cases/sample_info$total_n*100, 1)`%)

## Descriptive Statistics

```{r descriptive-stats}
# Filter to complete cases
analysis_data <- cleaned_data %>%
  filter(!is.na(public_trans_freq) & !is.na(prca_total) & !is.na(CAg) & !is.na(CAi))

descriptive_stats <- analysis_data %>%
  summarise(
    ptrs_mean = mean(public_trans_freq),
    ptrs_sd = sd(public_trans_freq),
    ptrs_min = min(public_trans_freq),
    ptrs_max = max(public_trans_freq),
    ca12_mean = mean(prca_total),
    ca12_sd = sd(prca_total),
    ca12_min = min(prca_total),
    ca12_max = max(prca_total),
    cag_mean = mean(CAg),
    cag_sd = sd(CAg),
    cai_mean = mean(CAi),
    cai_sd = sd(CAi)
  )
```

| Variable | Mean | SD | Range |
|----------|------|-----|-------|
| Public Transportation Frequency (PTRS) | `r round(descriptive_stats$ptrs_mean, 2)` | `r round(descriptive_stats$ptrs_sd, 2)` | `r descriptive_stats$ptrs_min` - `r descriptive_stats$ptrs_max` |
| Total Communication Apprehension (CA12) | `r round(descriptive_stats$ca12_mean, 2)` | `r round(descriptive_stats$ca12_sd, 2)` | `r descriptive_stats$ca12_min` - `r descriptive_stats$ca12_max` |
| Group Communication Apprehension (CAg) | `r round(descriptive_stats$cag_mean, 2)` | `r round(descriptive_stats$cag_sd, 2)` | 6 - 30 |
| Interpersonal Communication Apprehension (CAi) | `r round(descriptive_stats$cai_mean, 2)` | `r round(descriptive_stats$cai_sd, 2)` | 6 - 30 |

## Model Comparison

```{r model-comparison}
# Model 1: PTRS = b0 + b1*CA12 + e
model1 <- lm(public_trans_freq ~ prca_total, data = analysis_data)
model1_rmse <- sqrt(mean(residuals(model1)^2))
model1_r2 <- summary(model1)$r.squared

# Model 2: PTRS = b0 + b1*CAg + b2*CAi + e
model2 <- lm(public_trans_freq ~ CAg + CAi, data = analysis_data)
model2_rmse <- sqrt(mean(residuals(model2)^2))
model2_r2 <- summary(model2)$r.squared

# Model 3: PTRS = b0 + b1*CAg + b2*CAi + b3*(CAg * CAi) + e
analysis_data <- analysis_data %>%
  mutate(CAg_CAi_interaction = CAg * CAi)

model3 <- lm(public_trans_freq ~ CAg + CAi + CAg_CAi_interaction, data = analysis_data)
model3_rmse <- sqrt(mean(residuals(model3)^2))
model3_r2 <- summary(model3)$r.squared

# Create comparison table
model_comparison <- tibble(
  Model = c("Model 1: CA12 only", "Model 2: CAg + CAi", "Model 3: CAg + CAi + Interaction"),
  RMSE = c(model1_rmse, model2_rmse, model3_rmse),
  R_squared = c(model1_r2, model2_r2, model3_r2)
) %>%
  arrange(RMSE)
```

### RMSE Comparison

| Model | RMSE | R² |
|-------|------|-----|
| `r model_comparison$Model[1]` | `r round(model_comparison$RMSE[1], 4)` | `r round(model_comparison$R_squared[1], 4)` |
| `r model_comparison$Model[2]` | `r round(model_comparison$RMSE[2], 4)` | `r round(model_comparison$R_squared[2], 4)` |
| `r model_comparison$Model[3]` | `r round(model_comparison$RMSE[3], 4)` | `r round(model_comparison$R_squared[3], 4)` |

**Best Model:** `r model_comparison$Model[1]` with RMSE = `r round(model_comparison$RMSE[1], 4)`

## Model Details

### Model 1: Simple Regression (CA12 only)

```{r model1-details}
model1_tidy <- broom::tidy(model1)
model1_glance <- broom::glance(model1)

# Function to add significance stars
add_stars <- function(p) {
  stars <- ifelse(p < 0.001, "***",
          ifelse(p < 0.01, "**",
          ifelse(p < 0.05, "*", "")))
  return(stars)
}
```

| Predictor | Estimate | SE | t-value | p-value |
|-----------|----------|-----|---------|---------|
| Intercept | `r round(model1_tidy$estimate[1], 3)` | `r round(model1_tidy$std.error[1], 3)` | `r round(model1_tidy$statistic[1], 2)` | `r format(model1_tidy$p.value[1], scientific = TRUE, digits = 4)` `r add_stars(model1_tidy$p.value[1])` |
| CA12 | `r round(model1_tidy$estimate[2], 3)` | `r round(model1_tidy$std.error[2], 3)` | `r round(model1_tidy$statistic[2], 2)` | `r format(model1_tidy$p.value[2], scientific = TRUE, digits = 4)` `r add_stars(model1_tidy$p.value[2])` |

**Model Fit:** R² = `r round(model1_r2, 4)`, RMSE = `r round(model1_rmse, 4)`

### Model 2: Multiple Regression (CAg + CAi)

```{r model2-details}
model2_tidy <- broom::tidy(model2)
model2_glance <- broom::glance(model2)
```

| Predictor | Estimate | SE | t-value | p-value |
|-----------|----------|-----|---------|---------|
| Intercept | `r round(model2_tidy$estimate[1], 3)` | `r round(model2_tidy$std.error[1], 3)` | `r round(model2_tidy$statistic[1], 2)` | `r format(model2_tidy$p.value[1], scientific = TRUE, digits = 4)` `r add_stars(model2_tidy$p.value[1])` |
| CAg | `r round(model2_tidy$estimate[2], 3)` | `r round(model2_tidy$std.error[2], 3)` | `r round(model2_tidy$statistic[2], 2)` | `r format(model2_tidy$p.value[2], scientific = TRUE, digits = 4)` `r add_stars(model2_tidy$p.value[2])` |
| CAi | `r round(model2_tidy$estimate[3], 3)` | `r round(model2_tidy$std.error[3], 3)` | `r round(model2_tidy$statistic[3], 2)` | `r format(model2_tidy$p.value[3], scientific = TRUE, digits = 4)` `r add_stars(model2_tidy$p.value[3])` |

**Model Fit:** R² = `r round(model2_r2, 4)`, RMSE = `r round(model2_rmse, 4)`

### Model 3: Multiple Regression with Interaction (CAg + CAi + Interaction)

```{r model3-details}
model3_tidy <- broom::tidy(model3)
model3_glance <- broom::glance(model3)
```

| Predictor | Estimate | SE | t-value | p-value |
|-----------|----------|-----|---------|---------|
| Intercept | `r round(model3_tidy$estimate[1], 3)` | `r round(model3_tidy$std.error[1], 3)` | `r round(model3_tidy$statistic[1], 2)` | `r format(model3_tidy$p.value[1], scientific = TRUE, digits = 4)` `r add_stars(model3_tidy$p.value[1])` |
| CAg | `r round(model3_tidy$estimate[2], 3)` | `r round(model3_tidy$std.error[2], 3)` | `r round(model3_tidy$statistic[2], 2)` | `r format(model3_tidy$p.value[2], scientific = TRUE, digits = 4)` `r add_stars(model3_tidy$p.value[2])` |
| CAi | `r round(model3_tidy$estimate[3], 3)` | `r round(model3_tidy$std.error[3], 3)` | `r round(model3_tidy$statistic[3], 2)` | `r format(model3_tidy$p.value[3], scientific = TRUE, digits = 4)` `r add_stars(model3_tidy$p.value[3])` |
| CAg × CAi | `r round(model3_tidy$estimate[4], 3)` | `r round(model3_tidy$std.error[4], 3)` | `r round(model3_tidy$statistic[4], 2)` | `r format(model3_tidy$p.value[4], scientific = TRUE, digits = 4)` `r add_stars(model3_tidy$p.value[4])` |

**Model Fit:** R² = `r round(model3_r2, 4)`, RMSE = `r round(model3_rmse, 4)`

## Visualization

```{r visualization}
# Create a comparison plot
rmse_data <- model_comparison %>%
  mutate(Model = factor(Model, levels = Model))

ggplot(rmse_data, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_text(aes(label = round(RMSE, 4)), vjust = -0.5) +
  labs(
    title = "RMSE Comparison Across Models",
    x = "Model",
    y = "RMSE (Lower is Better)",
    fill = "Model"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## Summary

Analysis of `r sample_info$complete_cases` complete cases reveals that **`r model_comparison$Model[1]`** provides the best prediction of public transportation use frequency with the lowest RMSE of `r round(model_comparison$RMSE[1], 4)`. 

- **Model 1 (CA12 only):** RMSE = `r round(model1_rmse, 4)`, R² = `r round(model1_r2, 4)`
- **Model 2 (CAg + CAi):** RMSE = `r round(model2_rmse, 4)`, R² = `r round(model2_r2, 4)`  
- **Model 3 (CAg + CAi + Interaction):** RMSE = `r round(model3_rmse, 4)`, R² = `r round(model3_r2, 4)`

The results suggest that using the total communication apprehension score (CA12) provides the most accurate prediction of public transportation use frequency compared to separating group and interpersonal apprehension or including their interaction.

# What questions or uncertainties remain?

- **Model Complexity:** Does the simpler Model 1 generalize better to new data despite having similar predictive performance?
- **Theoretical Interpretation:** What does the lack of improvement from separating CAg and CAi suggest about the relationship between communication apprehension dimensions and transportation behavior?
- **Interaction Effects:** Why does the interaction term not improve prediction? Is there a non-linear relationship that wasn't captured?
- **Alternative Specifications:** Would other model specifications (e.g., polynomial terms, different transformations) improve prediction accuracy?
